Measuring contrast processing in the visual system using the steady state visually evoked potential (SSVEP)

### 2024-12-03 review.qmd
Propagated changes to R version

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. Because they can arise from different neuronal computations, different populations and even different stages of the visual system, different harmonics can exhibit different input-output functions. For example, 1F and 2F responses can have different thresholds [@BobakIOVS1984] and scalp topographies [@reganEvokedPotentialsSpecific1973] while Kaester et al [@KaestnerTVST2024] show both different response slopes and thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@reganEvokedPotentialsSpecific1973; @reganColourCodingPattern1975; @reganEffectStimulusColour1966a; @McKeefry1996; @Di-Russo2001] (see also [@Baseler1997]).

@@

These contrast definitions are appropriate both to photometric measures of stimulus contrast [for example, luminance; @Lennie1993] and also to definitions based on cone excitations [@Derrington1984; @MacLeod1979] which are more common in work on chromatic processing.

@@

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@reganRapidObjectiveRefraction1973; @Tyler1979]. To avoid hysteresis or short-term adaptation effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

The approach of estimating visual thresholds by extrapolating SSVEP responses at higher stimulus levels has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity and acuity for both achromatic and chromatic stimuli as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b]. For example, using SSVEPs, Norcia and Tyler [@Norcia1985b] found acuity thresholds in one month old infants that were roughly half those estimated previously from behavioural experiments (reviewed in [@brownDevelopmentVisualSensitivity1990a]). The ability to make very sensitive measurements of early visual function in infants also allows researchers to monitor the trajectory of visual development and identify different temporal epochs of developmental change [@Norcia1990].

SSVEP measures of thresholds for detecting purely chromatic stimuli find responses to isoluminant stimuli by around 5 to six weeks with contrast detection reaching near-adult levels by around six months [@morroneDevelopmentInfantContrast1993]. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around one year [@Norcia1985b; @Hamer1989] compared to around six to seven years with behavioural measures [@Ellemberg1999; @Atkinson1983].

At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which require careful measurement of the infant's gaze direction.  In slightly older chilren, other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section. Although response- [@Morrone1986] or contrast gain control [@Candy2001; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

@@

Amblyopia is a disorder of binocular vision, characterised by one eye contributing much less to perception than the other. This is often due to strabismus (squint) or anisometropia (difference in optical prescription between the eyes) during development. Contemporary accounts suggest that the amblyopic eye is suppressed by signals from the fellow eye. SSVEPs provide a convenient and objective method to characterise the difference in neural response to signals in each eye, and typically show reduced responses to stimuli in the amblyopic eye (see [@fig-plotclinical]b) across the contrast range [@Baker2015; @Lygo2021]. Measurments of interocular contrast modulation (for example, [@norciaChangesCorticalActivity2000; @Hou-houchuan2021] can also provide information about the way that amblyopia changes the balance of excitation and inhibition between eyes.

There are currently many novel binocular treatments for amblyopia under development, often involving virtual reality or stereo display systems designed to encourage the two eyes to work together. The steady-state approach may be more sensitive and objective than typical acuity measurements, and also has the potential to measure suppression between the eyes directly [e.g. @Zheng2019; @Du2023; @Hu2023].

@@

Later in the visual pathway, we would like to know more about the role of corticothalamic feedback in the LGN. Although it is commonly thought of as a simple relay between the eye and the cortex,  the majority of inputs to the LGN come from cortex not the eye, and there is good evidence that contrast processing in the LGN can be altered by top-down signals including attention [@Briggs2007; @Sherman1996; @OConnor2002; @Gouws2014]. Bottom-up inputs to the LGN are segregated by eye and so responses there are often considered to be purely monocular, but it is possible that feedback signals allow some binocular computations such as interocular normalisation [@Baker2007; @Dougherty2019] to begin even at this relatively early stage. Frequency tagging inputs to different eyes or in different precortical pathways allows us to address neurons in different parts of the LGN. These techniques combined with advances in recording technology (for example, sensitive source-imaged EEG and MEG recordings that can resolve subcortical structures [@Attal2012; @Attal2013; @Tesche1996], noninvasive deep-brain stimulation techniques [@Mohammadjavadi2022] or implanted electrode arrays [@Krolak-Salmon2003]) may allow us to study these computations in more detail.

Finally, SSVEPs continue to provide a way of studying contrast in the cortex. Here, we are often interested in how different visual parameters interact. For example, how are signals from different eyes combined to generate both scalar contrast values and also binocular depth cues? How does contrast combination depend on the low level properties of the individual inputs such as their retinotopic location, cone contrast, spatiotemporal frequency, eye of origin and orientation? Colour vision scientists are particularly interested in how chromatic signals originating in a small number of cone-opponent retinal pathways are transformed into a perceptual colour space where the 'unique hues' appear to be only weakly-related to the early retinal outputs and how these computations are conditioned by the spatial and temporal properties of the scene [@Li2022; @Stoughton2008; @Kaneko2020; @Gegenfurtner2003; @Solomon2007; @Wandell1993]. All of these questions can be addressed by using SSVEP and frequency tagging to examine the computations that combine and transform the inputs at different cortical stages [@Watts2024; @Baker2017; @Chen2021; @Katyal2018; @Busse2009]. The original promise of the SSVEP approach was that the  entire complex-valued frequency spectrum recorded at each location provided detailed information about the processing nonlinearities up to and including that point. In principle, this would allow researchers to uniquely fit the parameters of their computational models of visual processing [@Regan1988].  To date, the complexity of the neuronal computations at even the earliest stage of visual processing have hampered this effort - researchers typically restrict their analyses to the amplitudes of single, low-order frequency components (for example, the input frequencies or simple sums and differences of those frequencies). As our understanding of the early visual system improves, it is becoming possible to generate more realistic parameterised forward models of signal generation [@Chariker2020; @Groen2022; @Baker2017; @Tsai2012; @Schrimpf2020]. Feeding frequency-tagged inputs into these types of model allows us to generate synthetic SSVEP responses that can be compared with those measured in human subjects. In principle, we are therefore able to use the SSVEP to fit the parameters of early visual processing. This approach may also allow us to develop more sensitive tests for the changes in early visual processing that accompany a wide range of neurological diseases and disorders.

>>>>

The higher harmonics of the steady-state signal are generally thought to reflect nonlinear processing in the visual system [@Regan1988]. Because they can arise from different neuronal computations, different populations and even different stages of the visual system, different harmonics can exhibit different input-output functions. For example, 1F and 2F responses can have different thresholds [@BobakIOVS1984] and scalp topographies [@reganEvokedPotentialsSpecific1973] while @KaestnerTVST2024 show both different response slopes and thresholds for 1F and 2F components generated by dynamic noise.

@@

Contrast is one of the most fundamental pieces of information that the eye transmits to the brain. It can be defined as the change in cone photoreceptor activity over space ('spatial contrast') or time ('temporal contrast'). Cone photoreceptors - which drive precortical opponent pathways -  contribute to both chromatic and achromatic contrast, and although most of the research we describe here focuses on achromatic contrast, SSVEPs have proven to be an excellent measure of early chromatic processing as well [@reganEvokedPotentialsSpecific1973; @reganColourCodingPattern1975; @reganEffectStimulusColour1966a; @McKeefry1996; @Di-Russo2001] [see also @Baseler1997].

@@

These contrast definitions are appropriate both to photometric measures of stimulus contrast (for example, luminance; @Lennie1993) and also to definitions based on cone excitations [@Derrington1984; @MacLeod1979] which are more common in work on chromatic processing.

@@

A robust estimate of the threshold therefore requires the measurement of the SSVEP amplitude at many different super-threshold contrast levels. This was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@reganRapidObjectiveRefraction1973; @Tyler1979]. To avoid hysteresis or short-term adaptation effects, the sweep is sometimes conducted both up and down in the same experiment [@Tyler1979; @Norcia1985a; @Norcia1990]. The sweep VEP (really, a 'sweep SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004; @Hoffmann2017; @Bach2008; @Bach2020].

The approach of estimating visual thresholds by extrapolating SSVEP responses at higher stimulus levels has revealed much about the development of visual abilities in infants [@Harris1976; @Braddick1986; @Atkinson1979]. In general, SSVEP measurements of infant vision have revealed that contrast sensitivity and acuity for both achromatic and chromatic stimuli develops earlier than had been supposed previously based on behavioural readouts [@Dobson1978; @Norcia1985b]. For example, using SSVEPs, @Norcia1985b found acuity thresholds in one month old infants that were roughly half those estimated previously from behavioural experiments [reviewed in @brownDevelopmentVisualSensitivity1990a]. The ability to make very sensitive measurements of early visual function in infants also allows researchers to monitor the trajectory of visual development and identify different temporal epochs of developmental change [@Norcia1990].

SSVEP measures of thresholds for detecting purely chromatic stimuli find responses to isoluminant stimuli by around 5-6 weeks postnatally, with contrast detection reaching near-adult levels by around six months [@morroneDevelopmentInfantContrast1993]. Spatial acuity as measured by SSVEP reaches adult levels more slowly, but near-adult levels are recorded around one year [@Norcia1985b; @Hamer1989] compared to around six to seven years with behavioural measures [@Ellemberg1999; @Atkinson1983].

At least some of this difference is likely due to the relative objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking, which requires careful measurement of the infant's gaze direction. In slightly older chilren, other groups have reported electrophysiological correlates of visual acuity that more closely match the behavioural measures [@De-Vries-Khoe1982].

The SSVEP technique has also been used to study the development of the contrast gain control mechanisms described in the previous section. Although response gain [@Morrone1986] or contrast gain control [@Candy2001; @Skoczenski1998] is measurable in infants as young as six weeks old, its development appears to be slower, with adult levels being reached at approximately 11 years [@Pei2017].

@@

Amblyopia is a disorder of binocular vision, characterised by one eye contributing much less to perception than the other. This is often due to strabismus (squint) or anisometropia (difference in optical prescription between the eyes) during development. Contemporary accounts suggest that the amblyopic eye is suppressed by signals from the fellow eye. SSVEPs provide a convenient and objective method to characterise the difference in neural response to signals in each eye, and typically show reduced responses to stimuli in the amblyopic eye (see [@fig-plotclinical]b) across the contrast range [@Baker2015; @Lygo2021]. Measurements of interocular contrast modulation [for example, @norciaChangesCorticalActivity2000; @Hou-houchuan2021] can also provide information about the way that amblyopia changes the balance of excitation and inhibition between eyes. There are currently many novel binocular treatments for amblyopia under development, often involving virtual reality or stereo display systems designed to encourage the two eyes to work together. The steady-state approach may be more sensitive and objective than typical acuity measurements, and also has the potential to measure suppression between the eyes directly [e.g. @Zheng2019; @Du2023; @Hu2023].

@@

Later in the visual pathway, we would like to know more about the role of corticothalamic feedback in the LGN. Although it is commonly thought of as a simple relay between the eye and the cortex, the majority of inputs to the LGN come from cortex rather than the eye, and there is good evidence that contrast processing in the LGN can be altered by top-down signals including attention [@Briggs2007; @Sherman1996; @OConnor2002; @Gouws2014]. Bottom-up inputs to the LGN are segregated by eye and so responses there are often considered to be purely monocular, but it is possible that feedback signals allow some binocular computations such as interocular normalisation [@Baker2007; @Dougherty2019] to begin even at this relatively early stage. Frequency tagging inputs to different eyes or in different precortical pathways allows us to address neurons in different parts of the LGN. These techniques combined with advances in recording technology (for example, sensitive source-imaged EEG and MEG recordings that can resolve subcortical structures [@Attal2012; @Attal2013; @Tesche1996], noninvasive deep-brain stimulation techniques [@Mohammadjavadi2022] or implanted electrode arrays [@Krolak-Salmon2003]) may allow us to study these computations in more detail.

Finally, SSVEPs continue to provide a way of studying contrast in the cortex. Here, we are often interested in how different visual parameters interact. For example, how are signals from different eyes combined to generate both scalar contrast values and also binocular depth cues? How does contrast combination depend on the low level properties of the individual inputs such as their retinotopic location, cone contrast, spatiotemporal frequency, eye of origin and orientation? Colour vision scientists are particularly interested in how chromatic signals originating in a small number of cone-opponent retinal pathways are transformed into a perceptual colour space where the 'unique hues' appear to be only weakly-related to the early retinal outputs and how these computations are conditioned by the spatial and temporal properties of the scene [@Li2022; @Stoughton2008; @Kaneko2020; @Gegenfurtner2003; @Solomon2007; @Wandell1993]. All of these questions can be addressed by using SSVEP and frequency tagging to examine the computations that combine and transform the inputs at different cortical stages [@Watts2024; @Baker2017; @Chen2021; @Katyal2018; @Busse2009].

The original promise of the SSVEP approach was that the  entire complex-valued frequency spectrum recorded at each location provided detailed information about the processing nonlinearities up to and including that point. In principle, this would allow researchers to uniquely fit the parameters of their computational models of visual processing [@Regan1988]. To date, the complexity of the neuronal computations at even the earliest stage of visual processing have hampered this effort - researchers typically restrict their analyses to the amplitudes of single, low-order frequency components (for example, the input frequencies or simple sums and differences of those frequencies). As our understanding of the early visual system improves, it is becoming possible to generate more realistic parameterised forward models of signal generation [@Chariker2020; @Groen2022; @Baker2017; @Tsai2012; @Schrimpf2020]. Feeding frequency-tagged inputs into these types of model allows us to generate synthetic SSVEP responses that can be compared with those measured in human subjects. In principle, we are therefore able to use the SSVEP to derive the parameters of early visual processing. This approach may also allow us to develop more sensitive tests for the changes in early visual processing that accompany a wide range of neurological diseases and disorders.
